Recognizing Strong Gravitational Lenses

### 2015-03-16 doc/Course_Paper/paper.tex
more paper

A consequence of Einstein's Theory of General Relativity is that mass bends the
path of light. \cite{Einstein:2007aa} Most of the time the deflections are very
small; the original 'gravitational lens' that tested the veracity of Einstein's
theory in the years after the first world war was the sun, which deflected
light from stars behind it only a few seconds of arc. However, when light
passes through a particularly deep gravitational potential (say, near the
center of the dark matter halo of a galaxy cluster), the deflections can be
particularly large, resulting in brilliant arcs and multiple images. These
strong deflections due to light passing through a deep gravitational potential
are termed strong gravitational lenses. The gravitational lensing signal can
heuristically be thought of as a trade off between a couple factors: larger
gradients in the gravitational potential create larger distortions (and larger
gradients in the gravitational potential tend to reside near to the center of
the foreground galaxy or galaxy cluster); the more separated the foreground and
background objects are, the bigger the proportional distortion on the
background object by the foreground (but the farther away the background object
is, the smaller it appears\footnote{Note that this is only true in the ``low
  redshift'' universe: when objects are farther away than a cosmological
redshift of $z \approx 2.5$ or about 2.5 Gyr after the birth of the Universe,
they will actually \textit{grow} in angular extent.}). \cite{Kochanek:2006oz}

@@

From approximately 12000 fields of $440\times440\times3$ fields, we have
constructed approximately 30000 cutouts sized $96\times96\times3$. These cutouts
are selected based on where citizen-scientists clicked, on the theory that both
'correct' and 'incorrect' selections provide useful information about the
characteristics of gravitational lenses. In general, we have access to two
broad classes of images: 'training' and 'test' images. The 'training' images
include fields that were verified in advance to not contain any lenses as well
as simulated lensed galaxies, quasars, and clusters. Many of the simulated
objects are over-exaggerated and extremely obvious, but we also have access to
a second 'refinement stage' of the project, where much harder simulations were
given to users. The 'test' images are the fields that citizen-scientists
viewed, assessing whether a lens was in the field or not. In these 'test'
images are 120 known strong gravitational lens systems, which are also included
in this set. (The project confirms roughly half of these known lenses for
reasonable definitions of completeness and purity.) For all the images we also
have an associated probability that the project would evaluate that system as
containing a lens.

@@

% TODO: Insert figure for architecture

@@

% TODO: Touch up description
\begin{figure}%[!ht]
\centering
\epsfig{file=Figures/architecture.pdf,width=\linewidth,angle=0,clip=}
\caption{Architecture of the convolution
  neural network trained on the galaxy zoo morphologies. The feature vector we
use comes from the fully connected layer.}
\label{fig:ryan_architecture}
\end{figure}

@@

%-------------------------------------------------------------------------
\subsection*{Acknowledgements}
We received access to the dataset through Phil Marshall
\footnote{\url{pjm@slac.stanford.edu}}, who also graciously explained to us how
\textsc{Space Warps}
currently works. The convolutional neural network upon which the transfer
learning is based was kindly provided to us by Ryan Keisler
\footnote{\url{rkeisler@stanford.edu}}.
All errors in interpretation or otherwise are entirely our own.

>>>>

A consequence of Einstein's Theory of General Relativity is that mass bends the
path of light. \cite{Einstein:2007aa} Most of the time the deflections are very
small; the original 'gravitational lens' that tested the veracity of Einstein's
theory in the years after the first world war was the sun, which deflected
light from stars behind it only a few seconds of arc. However, when light
passes through a particularly deep gravitational potential (say, near the
center of the dark matter halo of a galaxy cluster), the deflections can be
particularly large, resulting in brilliant arcs and multiple images. These
strong deflections due to light passing through a deep gravitational potential
are termed strong gravitational lenses. The gravitational lensing signal can
heuristically be thought of as a trade off between a couple of factors: larger
gradients in the gravitational potential create larger distortions (and larger
gradients in the gravitational potential tend to reside near to the center of
the foreground galaxy or galaxy cluster), while the more separated the foreground and
background objects are, the bigger the proportional distortion on the
background object by the foreground (conversely, the farther away the background object
is, the smaller it appears\footnote{Note that this is only true in the ``low
  redshift'' universe: when objects are farther away than a cosmological
redshift of $z \approx 2.5$ or about 2.5 Gyr after the birth of the Universe,
they will actually \textit{grow} in angular extent.}). \cite{Kochanek:2006oz}

@@

From approximately 12000 fields of $440\times440\times3$ fields, we have
constructed approximately 30000 cutouts sized $96\times96\times3$. These
cutouts are selected based on where citizen-scientists clicked, on the theory
that both `correct' and `incorrect' selections provide useful information about
the characteristics of gravitational lenses. In general, we have access to two
broad classes of images: `training' and `test' images. The `training' images
include fields that were verified in advance to not contain any lenses as well
as simulated lensed galaxies, quasars, and clusters. Many of the simulated
objects are over-exaggerated and extremely obvious, but we also have access to
a second `refinement stage' of the project, where much harder simulations were
given to users. The `test' images are the fields that citizen-scientists
viewed, assessing whether a lens was in the field or not. In these 'test'
images are 120 known strong gravitational lens systems, which are also included
in this set. (The project confirms roughly half of these known lenses for
reasonable definitions of completeness and purity.) For all the images we also
have an associated probability that the project would evaluate that system as
containing a lens.

@@

\begin{figure}%[!ht]
\centering
\epsfig{file=Figures/theano_roc.pdf,width=\linewidth,angle=0,clip=}
\caption{Receiver Operating Curves extracted from the training and validation sets used with 
our Convolutional Neural Network after training for 35 epochs. These results should be compared 
with the \textsc{Space Warps} Stage 1 Receiver Operating Curve in Figure~\ref{fig:roc_tl}, which 
outperforms our network for false positive rates below $\sim 0.3$. However, above this 
point our Convolutional Net achieves higher true positive rates than any of the other (machine 
or human) methods herein considered.}
\label{fig:roc_theano}
\end{figure}

In order to retain our ability to experiment with non-traditional training methods and 
architectures, we chose to implement our Convolutional Neural Network (CNN) from the ground 
up using \textsc{theano}. We settled on a five-layer layer architecture consisting of two 
convolution/max dropout layers, a fully connected layer, and a softmax layer. Training was 
carried out using RMSprop with a decay rate of 0.9, and the model was regularized using $L_2$ normalization 
with a regularization strength of $10^{-4}$ and dropout with probability 0.5 at each layer . We chose our convolution layers to have stride 4 and 
depth 10 (with padding to maintain the input image size), and the subsequent max pooling layers to have stride 2. Assigning our fully 
connected layer 50 neurons (and including the paramters in the subsequent softmax layer) gave 
our model a total of 288,492 trainable parameters.

Due to memory and time constraints, we have not yet trained this CNN on all our of data. The results 
of training it on 8,000 randomly selected images from the \textsc{Space Warps} Stage 1 data set for 
35 epochs is shown in Figure~\ref{fig:roc_theano}, where we have plotted the Receiver Operating Curve 
obtained from our model's predictions on the training set and a test set composed of another (non-overlapping)
 randomly selected 8,000 images from the \textsc{Space Warps} Stage 1 data set. This result should be easily improved on by training on more of 
 our images, by adding data augmentation, and by increasing the complexity of our CNN architecture. However, it already shows that training 
 a CNN solely on the \textsc{Space Warps} data set is a viable option, given enough CPU power and time.

@@

\begin{figure}%[!ht]
\centering
\epsfig{file=Figures/architecture.pdf,width=\linewidth,angle=0,clip=}
\caption{Architecture of the convolution
  neural network trained on the galaxy zoo morphologies. The feature vector we
use comes from the fully connected layer.}
\label{fig:ryan_architecture}
\end{figure}

@@

%-------------------------------------------------------------------------
\subsection*{Acknowledgements}
We received access to the dataset through Phil Marshall\footnote{\url{pjm@slac.stanford.edu}}, who also graciously explained to us how
\textsc{Space Warps}
currently works. The convolutional neural network upon which the transfer
learning is based was kindly provided to us by Ryan Keisler\footnote{\url{rkeisler@stanford.edu}}.
All errors in interpretation or otherwise are entirely our own.
